---
title: "GGplot2 Part 2"
author: "Seun Odeyemi"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    toc: true
    toc_depth: 4
---

```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(error = TRUE, collapse = TRUE, comment = "#>")
```

## Load Libraries
```{r load libraries, include=TRUE, message=FALSE}
library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(skimr)
library(knitr)
library(kableExtra)
library(RColorBrewer)
```


## Exploring the mtcars dataset

```{r exploring mtcars}
mtcars_tbl <- as_tibble(mtcars)

glimpse(mtcars_tbl)

kable(head(mtcars_tbl)) %>% 
  kable_styling(bootstrap_options = "striped", full_width = F)
```

## Smoothing

```{r geom_smooth, fig.width=6, fig.height=6, fig.align='center'}

par(mfrow = c(2, 2))

# A scatter plot with LOESS smooth
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  geom_smooth(method = "loess")


# A scatter plot with an ordinary Least Squares linear model
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  geom_smooth(method = "lm")


# The previous plot, without CI ribbon
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  geom_smooth(method = "lm", se=FALSE)

# The previous plot, without points
ggplot(mtcars, aes(x = wt, y = mpg)) +
  stat_smooth(method = "lm", se=FALSE)
```

You can use either `stat_smooth()` or `geom_smooth()` to apply a linear model. Remember to always think about how the examples and concepts we discuss throughout the data viz courses can be applied to your own datasets!

## Grouping Variables

```{r grouping_vars, fig.width=6, fig.height=6, fig.align='center'}
# 1 - Define cyl as a factor variable
ggplot(mtcars, aes(x = wt, y = mpg, col = factor(cyl))) +
  geom_point() +
  stat_smooth(method = "lm", se = FALSE)

# 2 - Plot 1, plus another stat_smooth() containing a nested aes()
ggplot(mtcars, aes(x = wt, y = mpg, col = factor(cyl))) +
  geom_point() +
  stat_smooth(method = "lm", se = FALSE) +
  stat_smooth(method = "lm", se = FALSE, group = 1)
```

Good job! Notice that we can use multiple aesthetic layers, just like we can use multiple geom layers. Each aesthetic layer can be mapped onto a specific geom.

## Modifying stat_smooth (1)

```{r mod_stat_smooth_1, fig.width=6, fig.height=6, fig.align='center'}

par(mfrow = c(2, 2))

# Plot 1: change the LOESS span
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  # Add span below
  geom_smooth(se = FALSE, span = 0.7)

# Plot 2: Set the second stat_smooth() to use LOESS with a span of 0.7
ggplot(mtcars, aes(x = wt, y = mpg, col = factor(cyl))) +
  geom_point() +
  stat_smooth(method = "lm", se = FALSE) +
  # Change method and add span below
  stat_smooth(method = "loess", aes(group = 1),
              se = FALSE, col = "black", span = 0.7)

# Plot 3: Set col to "All", inside the aes layer of stat_smooth()
ggplot(mtcars, aes(x = wt, y = mpg, col = factor(cyl))) +
  geom_point() +
  stat_smooth(method = "lm", se = FALSE) +
  stat_smooth(method = "loess",
              # Add col inside aes()
              aes(group = 1, col = "All"),
              # Remove the col argument below
              se = FALSE, span = 0.7)

# Plot 4: Add scale_color_manual to change the colors
myColors <- c(brewer.pal(3, "Dark2"), "black")
ggplot(mtcars, aes(x = wt, y = mpg, col = factor(cyl))) +
  geom_point() +
  stat_smooth(method = "lm", se = FALSE, span = 0.7) +
  stat_smooth(method = "loess", 
              aes(group = 1, col="All"), 
              se = FALSE, span = 0.7) +
  # Add correct arguments to scale_color_manual
  scale_color_manual("Cylinders", values = myColors)
```

## Modifying stat_smooth (2)

```{r mod_stat_smooth_2, fig.width=6, fig.height=6, fig.align='center', eval = FALSE, echo=FALSE}
par(mfrow = c(2, 3))

# Plot 1: Jittered scatter plot, add a linear model (lm) smooth
ggplot(Vocab, aes(x = education, y = vocabulary)) +
  geom_jitter(alpha = 0.2) +
  stat_smooth(method = "lm", se = FALSE) # smooth

# Plot 2: points, colored by year
ggplot(Vocab, aes(x = education, y = vocabulary, col = year)) +
  geom_jitter(alpha = 0.2) 

# Plot 3: lm, colored by year
ggplot(Vocab, aes(x = education, y = vocabulary, col = factor(year))) +
  stat_smooth(method = "lm", se = FALSE) # smooth
  
# Plot 4: Set a color brewer palette
ggplot(Vocab, aes(x = education, y = vocabulary, col = factor(year))) +
  stat_smooth(method = "lm", se = FALSE) +  # smooth
  scale_color_brewer()  # colors

# Plot 5: Add the group aes, specify alpha and size
ggplot(Vocab, aes(x = education, y = vocabulary, col = year, group = factor(year))) +
  stat_smooth(method = "lm", se = FALSE, alpha = 0.6, size = 2) +
  scale_color_gradientn(colors = brewer.pal(9, "YlOrRd"))
```

## Calculating Statics

### Quantiles 

The previous example used the Vocab dataset and applied linear models describing vocabulary by education for different years. Here we'll continue with that example by using stat_quantile() to apply a quantile regression (method rq).

By default, the 1st, 2nd (i.e. median), and 3rd quartiles are modeled as a response to the predictor variable, in this case education. Specific quantiles can be specified with the quantiles argument.

If you want to specify many quantile and color according to year, then things get too busy. We'll explore ways of dealing with this in the next chapter.

```{r stat_quantiles, fig.width=6, fig.height=6, fig.align='center', eval = FALSE, echo=FALSE}

# Use stat_quantile instead of stat_smooth
ggplot(Vocab, aes(x = education, y = vocabulary, col = year, group = factor(year))) +
  stat_quantile(alpha = 0.6, size = 2) +
  scale_color_gradientn(colors = brewer.pal(9,"YlOrRd"))

# Set quantile to 0.5
ggplot(Vocab, aes(x = education, y = vocabulary, col = year, group = factor(year))) +
  stat_quantile(quantiles = 0.5, alpha = 0.6, size = 2) +
  scale_color_gradientn(colors = brewer.pal(9,"YlOrRd"))

```

Quick quantiles! Quantile regression is a great tool for getting a more detailed overview of a large dataset.

### Sum

```{r stat_sum, fig.width=6, fig.height=6, fig.align='center', eval=FALSE, echo=FALSE}

# Plot 1: Jittering only
p <- ggplot(Vocab, aes(x = education, y = vocabulary)) +
  geom_jitter(alpha = 0.2)

# Plot 2: Add stat_sum
p +
  stat_sum() # sum statistic

# Plot 3: Set size range
p +
  stat_sum() + # sum statistic
  scale_size(range = c(1, 10)) # set size scale
```

Good job! Remember, typically we'd draw our models on top of the dots, but in this case we didn't so that we could just keep recycling the p object.

### Mean & Standard Deviation

```{r calculating_stats, message=FALSE}
set.seed(123)

xx <- rnorm(100)

mean(xx)

mean(xx) + (sd(xx) * c(-1, 1))

library(Hmisc)
smean.sdl(xx, mult = 1)

# ggplot2
mean_sdl(xx, mult = 1)

# we can use this in ggplot2 by calling the fun.data function
p <- ggplot(iris, aes(x = Species, y = Sepal.Length)) 

p + stat_summary(fun.data = mean_sdl, fun.args = list(mult = 1))


```

By default the `stat_summary()` function uses the `geom_pointrange()`, which requires `y`, `ymin`, and `ymax` - the exact variables returned by `mean.sdl`. So everything works very well together. If we wanted a more typical errorbar style plot, we can independently plot the mean and the use the `point` argument for the `geom` and again call `mean_sdl`, but this time using the errorbar geom. 

```{r stat_summary, fig.width=6, fig.height=6, fig.align='center'}
ggplot(iris, aes(x = Species, y = Sepal.Length)) + 
  stat_summary(fun.y = mean, geom = "point") + 
  stat_summary(fun.data = mean_sdl, 
                 fun.args = list(mult = 1), 
                 geom = "errorbar", width = 0.1)

ggplot(iris, aes(x = Species, y = Sepal.Length)) + 
  stat_summary(fun.y = mean, geom = "bar", fill = "skyblue") + 
  stat_summary(fun.data = mean_sdl, fun.args = list(mult = 1),
               geom = "errorbar", width = 0.1)

```

### Confidence Interval

```{r ninetyfivepercent_confidence_interval}
ERR <- qt(0.975, length(xx) - 1) * (sd(xx) / sqrt(length(xx)))

mean(xx) + (ERR* c(-1, 1))
            
smean.cl.normal(xx)

mean_cl_normal(xx)

ggplot(iris, aes(x = Species, y = Sepal.Length)) + 
  stat_summary(fun.data = mean_cl_normal, width = 0.1)
```

You can use any function in ggplot as long as the output has the expected format. Other useful `stat_` layer functions are `stat_summary`, `stat_function`, and `stat_qq`. 

| **stat**| **description** |
|:--------|:-------------|
|stat_summary()|summarize y values at distinct x values |
|stat_function()|compute y values from a function of x values|
|stat_qq()|perform calculations for a quantile-quantile plot|

### Normal distribution

```{r normal_distribution, fig.height=6, fig.width=6, fig.align='center'}
library(MASS)
mam.new <- data.frame(body = log10(mammals$body))
ggplot(mam.new, aes(x = body)) + 
  geom_histogram(aes( y = ..density..)) + 
  geom_rug() + 
  stat_function(fun = dnorm, colour = "red", 
                arg = list(mean = mean(mam.new$body),
                           sd = sd(mam.new$body)))

# QQplot

mam.new$slope <- diff(quantile(mam.new$body, c(0.25, 0.75))) /
diff(qnorm(c(0.25, 0.75)))
mam.new$int <- quantile(mam.new$body, 0.25) -
mam.new$slope * qnorm(0.25)
ggplot(mam.new, aes(sample = body)) + 
  stat_qq() + 
  geom_abline(aes(slope = slope, intercept = int), col = "red")
```

