---
title: "Probability and Distributions in R"
subtitle: "Multivariate Probability Distributions in R"
author: "Seun Odeyemi"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: united
    highlight: tango
    toc: true
    toc_depth: 4
    toc_float: true
---

```{r setup}
knitr::opts_chunk$set(error = TRUE, 
                      collapse = TRUE, 
                      comment = "#>")
set.seed(10^7*4)
library(styler)
library(lintr)
library(readr)
library(readxl)
library(here)
```


# Reading and plotting multivariate data

Multivariate distributions are designed to describe the probability distributions of more than one random variable at the same time. Since the variables are often correlated exploring them individually will only provide limited insight. 

In this course we will learn how to:

* Read and analyze multivariate data
* Explore plotting techniques
* Learn how to use common statistical distributions
  * Gaussian and T distribution
* Techniques for high-dimensional data
  * Principal component analysis (PCA)

Multivariate data is mostly rectangular in shape i.e. data is organized by rows and columns. Rows represent observations and columns represent variables. Data sets may or may not include row names/numbers or column headers. It is also important to be aware of the possibility of missing data. 

## Reading data into r

```{r iris_dataset}
iris_raw <- read_csv(here("datasets", "iris.txt"))
head(iris_raw)
```

## Assigning column names

```{r assign_column_names}
iris_raw <- iris_raw %>% 
  `colnames<-`(c("Sepal.Length", 
             "Sepal.Width", 
             "Petal.Length", 
             "Petal.Width",  
             "Species" )) 
head(iris_raw)

# Alternatively
# iris_raw2 <- read_csv(here("datasets", "iris.txt"))
# library(magrittr)
# iris_raw2 <- iris_raw2 %>%
#   set_colnames(c("Sepal.Length", 
#              "Sepal.Width", 
#              "Petal.Length", 
#              "Petal.Width",  
#              "Species" ))
# head(iris_raw2)

# OR
# colnames(iris_raw) <- c("Sepal.Length", 
#                         "Sepal.Width", 
#                         "Petal.Length", 
#                         "Petal.Width", 
#                         "Species" )
```


```{r birthweight_dataset}
birthweight <- read_csv(here("datasets", "birthweight.csv"))
str(birthweight)
head(birthweight)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
