---
title: "Probability and Distributions in R"
subtitle: "Multivariate Probability Distributions in R"
author: "Seun Odeyemi"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: united
    highlight: tango
    toc: true
    toc_depth: 4
    toc_float: true
---

```{r setup}
knitr::opts_chunk$set(error = TRUE, 
                      collapse = TRUE, 
                      comment = "#>")
set.seed(10^7*4)
library(styler)
library(lintr)
library(readr)
library(readxl)
library(here)
```


# Reading and plotting multivariate data

Multivariate distributions are designed to describe the probability distributions of more than one random variable at the same time. Since the variables are often correlated exploring them individually will only provide limited insight. 

In this course we will learn how to:

* Read and analyze multivariate data
* Explore plotting techniques
* Learn how to use common statistical distributions
  * Gaussian and T distribution
* Techniques for high-dimensional data
  * Principal component analysis (PCA)

Multivariate data is mostly rectangular in shape i.e. data is organized by rows and columns. Rows represent observations and columns represent variables. Data sets may or may not include row names/numbers or column headers. It is also important to be aware of the possibility of missing data. 

## Reading data into r

```{r iris_dataset}
iris_raw <- read_csv(here("datasets", "iris.txt"))
str(iris_raw)
```

## Assigning column names

```{r assign_column_names}
iris_raw <- iris_raw %>% 
  `colnames<-`(c("Sepal.Length", 
             "Sepal.Width", 
             "Petal.Length", 
             "Petal.Width",  
             "Species" )) 
head(iris_raw)

# Alternatively
# iris_raw2 <- read_csv(here("datasets", "iris.txt"))
# library(magrittr)
# iris_raw2 <- iris_raw2 %>%
#   set_colnames(c("Sepal.Length", 
#              "Sepal.Width", 
#              "Petal.Length", 
#              "Petal.Width",  
#              "Species" ))
# head(iris_raw2)

# OR
# colnames(iris_raw) <- c("Sepal.Length", 
#                         "Sepal.Width", 
#                         "Petal.Length", 
#                         "Petal.Width", 
#                         "Species" )
```

We need to change the last variable of the `iris` dataset, `Species`, to a categorical variable, which `R` calls a factor. Notice that `Species` is now a factor variable. 

```{r}
library(dplyr)
iris_raw <- iris_raw %>% mutate(Species = as.factor(Species))
glimpse(iris_raw)
```

We need to remove the unnecessary prefix "Iris-" in the Species column.

```{r}
iris_raw <- iris_raw %>% mutate(Species = gsub("Iris-", "", Species))
head(iris_raw)
```

# Practice 1

## 1: Reading multivariate data using read.table

Reading multivariate data from a file or an external URL can generally be done using the `read.table()` or `read.csv()` commands, depending on the type of data source. In this exercise you will read the wine dataset from the UCI machine learning website. A full description of the dataset can be found here.

```{r}
# Read in the wine dataset
wine <- read.table("http://archive.ics.uci.edu/ml/machine-learning-databases/wine/wine.data", sep = ",")

# Print the first four entries
head(wine)

# Find the dimensions of the data
dim(wine)

# Check the names of the wine dataset 
names(wine)
```

### Reading multivariate data in the tidyverse

```{r}
# Read in the wine dataset
wine_data <- read_delim("http://archive.ics.uci.edu/ml/machine-learning-databases/wine/wine.data", delim = ",")

# Print the first four entries
head(wine)

# Find the dimensions of the data
dim(wine)

# Check the names of the wine dataset 
names(wine)
```

## 2: Specifying datatypes for columns

When you read data from a text or CSV file, you should specify the names and data types for each column. The read() function will try to determine if the first entry of the dataset contains the column names. R is clever at figuring out some datatypes, but if you are reading a categorical variable coded as 0, 1, and 2, it will read it as a numeric variable, and you will need to specify the data type for that column after reading the data.

```{r}
# Assign new names
colnames(wine) <- c('Type', 'Alcohol', 'Malic', 'Ash', 'Alcalinity', 'Magnesium', 'Phenols', 'Flavanoids', 'Nonflavanoids','Proanthocyanins', 'Color', 'Hue', 'Dilution', 'Proline')
                      
# Check the new column names
names(wine)

# Check data type/structure of each variable
str(wine)

# Change the Type variable data type
wine$Type <- as.factor(wine$Type)

# Check data type/structure again 
str(wine)
```

### Specifying datatypes for columns in the tidyverse

```{r}
wine_data <- wine %>% `colnames<-`(c('Type', 'Alcohol', 'Malic', 
                            'Ash', 'Alcalinity', 'Magnesium', 
                            'Phenols', 'Flavanoids', 'Nonflavanoids',
                            'Proanthocyanins', 'Color', 'Hue', 'Dilution', 'Proline')) 
                      
# Check the new column names
names(wine)

# Check data type/structure of each variable
str(wine)

# Change the Type variable data type
library(forcats)
wine %>% mutate(Type = as_factor(Type))

# Check data type/structure again 
str(wine)
```

```{r}
library(skimr)
funs <- list(
  iqr = IQR,
  quantile = purrr::partial(quantile, probs = .99)
)

skim_with(numeric = funs, append = FALSE)
skim(iris, Sepal.Length)
```


```{r birthweight_dataset}
birthweight <- read_csv(here("datasets", "birthweight.csv"))
str(birthweight)
head(birthweight)

birthweight <- birthweight %>% select(-X1)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
