---
title: "Writing Functions in R (Chapter 2)"
subtitle: "When and How You Should Write a Function"
author: "Seun Odeyemi"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: united
    highlight: tango
    toc: true
    toc_depth: 4
    toc_float: true
---
```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(error = TRUE, 
                      collapse = TRUE, 
                      comment = "#>")
library(styler)
library(lintr)
```

# Why should you write a function?

Take a look at the code below. At a high level it looks as if we are repeating the same operation four times: we are transforming four columns in `df` in the same way. If you look closely, you'll notice that we take a column, substract the smallest value in the column and then divide by the range of that column. The goal of this code is to rescale the range of each column between (0, 1), but the repetition makes it hard to see that intent. It also makes it really hard to see mistakes. 

```{r motivating_a_need_for_fx}
df$a <- (df$a - min(df$a, na.rm = TRUE)) /
  (max(df$a, na.rm = TRUE) - min(df$a, na.rm = TRUE))
df$b <- (df$b - min(df$b, na.rm = TRUE)) /
  (max(df$a, na.rm = TRUE) - min(df$b, na.rm = TRUE))
df$c <- (df$c - min(df$c, na.rm = TRUE)) /
  (max(df$c, na.rm = TRUE) - min(df$c, na.rm = TRUE))
df$d <- (df$d - min(df$d, na.rm = TRUE)) /
  (max(df$d, na.rm = TRUE) - min(df$d, na.rm = TRUE))

```

## How was this code written?

1. I figured out how to rescale a column.
2. Then, once I was confident that I had gotten it right, I copied and pasted the code and then replaced each reference to column `a` with the appropriate column name (`b`, `c`, `d`). You can see how this can easily lead to mistakes. 

## How can you prevent this problem from happening?

This is Hadley Wickham's rule of thumb for when to stop copying-and-pasting and writing a function: as soon you've copied-and-pasted twice (so you've now got three copies of the code), it's time to write a function. There are other benefits to writing a function: even though we'll still repeat the same command four times, it's much easier to pick up mistakes since it takes less time to understand the intent of the code, we can spend more effort checking the correctness. Also, if we decide later that we want to do a different sort of operation, it's clear what we need to change. You might notice that there is still a lot of duplication here. Later in the course you'll learn how to do functional programming with the `purrr` package to reduce this even further. 

```{r create_a_sample_df}
set.seed(1223)
df <- data.frame(
  a = rnorm(10, mean = 5, sd = 1),
  b = rnorm(10, mean = 5, sd = 1),
  c = rnorm(10, mean = 5, sd = 1),
  d = rnorm(10, mean = 5, sd = 1)
)
# glimpse(df)
```

## Start with a snippet of code

We have a snippet of code that successfully rescales a column to be between 0 and 1:

>(df$a - min(df$a, na.rm = TRUE)) /  
  (max(df$a, na.rm = TRUE) - min(df$a, na.rm = TRUE))

Our goal over the next few exercises is to turn this snippet, written to work on the `a` column in the data frame `df`, into a general purpose `rescale01()` function that we can apply to any vector.

The first step of turning a snippet into a function is to examine the snippet and decide how many inputs there are, then rewrite the snippet to refer to these inputs using temporary names. These inputs will become the arguments to our function, so choosing good names for them is important. (We'll talk more about naming arguments in a later exercise.)

In this snippet, there is one input: the numeric vector to be rescaled (currently `df$a`). What would be a good name for this input? It's quite common in R to refer to a vector of data simply as x (like in the [mean](http://www.rdocumentation.org/packages/base/functions/mean) function), so we will follow that convention here.

```{r wrong_approach_to_fx_def, echo=FALSE}
rescale01 <- function(x) {
  (x - rng[1]) / 
    (rng[2] - rng[1])
}
df$a <- rescale01(df$a)
```

```{r def_a_sample_vector}
# Define example vector x
x <- c(1:10, NA)
# Rewrite this snippet to refer to x
(x - min(x, na.rm = TRUE)) /
  (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))
```

## Rewrite for clarity

Our next step is to examine our snippet and see if we can write it more clearly.

Take a close look at our rewritten snippet. Do you see any duplication?

>(x - min(x, na.rm = TRUE)) /
  (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))

One obviously duplicated statement is `min(x, na.rm = TRUE)`. It makes more sense for us just to calculate it once, store the result, and then refer to it when needed. In fact, since we also need the maximum value of `x`, it would be even better to calculate the _range_ once, then refer to the first and second elements when they are needed.

What should we call this intermediate variable? You'll soon get the message that using good names is an important part of writing clear code! I suggest we call it `rng` (for "range").

```{r rewrite_fx_for_clarity}
# Define example vector x
x <- c(1:10, NA)

# Define rng
rng <- range(x, na.rm = TRUE)

# Rewrite this snippet to refer to the elements of rng
(x - rng[1]) / 
  (rng[2] - rng[1])
```

## Finally turn it into a function

What do you need to write a function? You need a **name** for the function, you need to know the **arguments** to the function, and you need code that forms the **body** of the function.

We now have all these pieces ready to put together. It's time to write the function!

```{r turn_it_into_fx}
# Define example vector x
x <- c(1:10, NA) 

# Use the function template to create the rescale01 function
rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}
# Test your function, call rescale01 using the vector x as the argument
rescale01(x)
```

# How should you write a function?

What we just worked through is a pattern for exactly how we think you should approach writing a function. Perhaps most importantly, we didn't start by trying to fill out the function snippet. Instead we started with a concrete problem to solve. 

Here is a dataframe `df`. We want to scale column `a` to be between 0 and 1. 

```{r create_df}
set.seed(1223)
df <- data.frame(
  a = rnorm(10, mean = 5, sd = 1),
  b = rnorm(10, mean = 5, sd = 1),
  c = rnorm(10, mean = 5, sd = 1),
  d = rnorm(10, mean = 5, sd = 1)
)
glimpse(df)
```

1. Start with a simple problem

Start with a column that wasn't random. That way we know what the answer should be. 

```{r how_to_write_a_fx_1}
df <- data.frame(
  a = 1:10,
  b = rnorm(10, mean = 5, sd = 1),
  c = rnorm(10, mean = 5, sd = 1),
  d = rnorm(10, mean = 5, sd = 1)
)
# Output should be:
seq(from = 0.0, to = 1.0, by = 0.1)
```

2. Get a working snippet of code

Get a working snippet of code that solves our simple problem. 

```{r how_to_write_a_fx_2}
(df$a - min(df$a, na.rm = TRUE)) /  
  (max(df$a, na.rm = TRUE) - min(df$a, na.rm = TRUE))
```

3. Rewrite to use temporary variables

To turn a snippet into a function, we examine what the snippet should be then we rewrite the snippet to refer to these inputs using temporary variables. It can be useful at this point to rewrite our problem in terms of these variables to help test our snippet. 

> (x - min(x, na.rm = TRUE)) /
  (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))

4. Rewrite for clarity

We then a take careful look at our snippet and decide if we can write it in a clearer way. In our example we can make our intention clearer by extracting the calculation of the range. 

```{r how_to_write_a_fx_4}
x <- df$a

rng <- range(x, na.rm = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
```


5. Turn it into a function

Only now, once we have a clearly-written snippet do we turn into a function

```{r how_to_write_a_fx_5}
x <- df$a

# Use the function template to create the rescale01 function
rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}
# Test your function, call rescale01 using the vector x as the argument
rescale01(x)
```

# Practice 1

1. Start with a simple problem

```{r simple_problem}
# Define example vectors x and y
x <- c( 1, 2, NA, 3, NA)
y <- c(NA, 3, NA, 3,  4)

# Count how many elements are missing in both x and y
sum(is.na(x) & is.na(y)) # or sum(is.na(x && y))
```

2. Rewrite snippet as function

```{r rewrite_snippet_as_fx}
# Define example vectors x and y
x <- c( 1, 2, NA, 3, NA)
y <- c(NA, 3, NA, 3,  4)
# Turn this snippet into a function: both_na()
both_na <- function(x, y){
  sum(is.na(x) & is.na(y))
}
```


3. Put our function to use

```{r put_fx_to_use}
# Define x, y1 and y2
x <-  c(NA, NA, NA)
y1 <- c( 1, NA, NA)
y2 <- c( 1, NA, NA, NA)
# Call both_na on x, y1
both_na(x, y1)
# Call both_na on x, y2
both_na(x, y2)
```

Nice! Are the answers what you expected? What should both_na(x, y2) return? You might argue it should return an error. You'll see how to handle this in a later chapter. 

# How to write a good function? 

What maskes a good function? 

1. Correctness: The first rule in creating a function is that said function solves a problem **correctly**. 
2. Understandable: A good function must be understandable not only to R but to humans. Note: Functions are for humans and computers. The combination of being correct and understandable is powerful. It leads to obviously correct functions i.e. functions that are easily verifiable to humans to do exactly what they set out to do. 

What does this code do?

```{r guess_what_this_does, eval=FALSE}
bar <- foo(bar, qux)
```

You are probably have a hard time figuring it out. What does the function `foo` do? What are the objects `bar` and `qux` doing? In isolation this line of code is meaningless. To figure it out we need to search back to earlier code to understand what these objects are. 

Check out the line of code below

```{r this_is_better, eval=FALSE}
df2 <- arrange(df, qux)
```

It is surprisingly similar to the previous code, but some subtle clues help us figure out what might be happening. We can probably guess that we're creating a new dataframe by arranging the dataframe `df` according to something called `qux`. 

> Just using good names helps us understand code with minimal context. 

# Naming principles

Note: This applies to objects, functions, or arguments. 

1. Pick a consistent style for long names. 

It is best to always use lower case and separating words in long names with an `_`. There are other conventions but it doesn't matter which convention you use but stick to it. Don't mix conventions in your code. 

```{r use_good_names, eval=FALSE}
# Good
col_mins()
row_maxes()
# Bad
newData <- c(old.data, todays_log)
```

2. Do not override existing variables or functions. 

  * Don't define capital `T` or `F` because that can be confused with shorthand for `TRUE` or `FALSE`. 
  * Don't call an object `c` or a new function `mean` since that can be confused for existing functions in R.  

```{r donot_override_variables, eval=FALSE}
# Bad
T <- FALSE
c <- 10
mean <- function(x) sum(x)
```

3. When it comes to naming functions you should generally use **verbs** i.e. doing words since functions do things. For example, a function that imputes missing values might be called `impute_missing()` instead of `imputed()`. 

4. Functions: Make sure your name describes what the function does. No matter how awesome your function is, `my_awesome_function()` is a bad name. 

5. Arguments: You should follow the same good naming rules for arguments except arguments should be nouns i.e. naming words instead of verbs. 

  a. Use common short names when appropriate:
    * vectors can be referred to as `x`, `y`, or `z`.
    * `df` usually refers to a dataframe.
    * `i`, `j` are normally reserved for indices (typically rows and columns).
    * `n`: length or number of rows
    * `p`: number of columns
  b. The order of the argument is important too. Arguments generally fall into two types: 1. Data arguments used to supply data to be computed on like `x` in the mean function and `x` and `y` in the `t.test` function below; 2. Detail arguments supply arguments that control the details of the computation. The mean function has an argument `na.rm` for determining how missing values should be treated. The t.test function has a whole host of arguments controlling the alternative hypothesis to be tested, the type of test, and the confidence level. You should always put data arguments first followed by detailed arguments. Detailed arguments should always be given default values.

```{r order_of_arguments, eval=FALSE}
mean(x, trim = 0, na.rm = FALSE, ...)

t.test(x,
  y = NULL,
  alternative = c("two.sided", "less", "greater"),
  mu = 0, paired = FALSE, var.equal = FALSE,
  conf.level = 0.95, ...
)
```

6. Make it clear what the function returns. 

# Practice II

1. Good function names

Consider the following function, `f2()`, which has also been loaded into your workspace:

```r 
f2 <- function(x) {
  if (length(x) <= 1) return(NULL)
  x[-length(x)]
}
```
`f2()` isn't a very good function name! Which of the following would make a good name for this function? Answer: `remove_last()`

2. Argument names

```{r}
# Rewrite mean_ci to take arguments named level and x
mean_ci <- function(level, x) {
  se <- sd(x) / sqrt(length(x))
  alpha <- 1 - level
  half_alpha <- alpha / 2
  mean(x) + se * qnorm(c(half_alpha, 1 - half_alpha))
}
```

3. Argument order

Aside from giving your arguments good names, you should put some thought into what order your arguments are in and if they should have defaults or not.

Arguments are often one of two types:

* Data arguments supply the data to compute on.
* Detail arguments control the details of how the computation is done.

Generally, data arguments should come first. Detail arguments should go on the end, and usually should have default values. 

```{r}
# Alter the arguments to mean_ci
mean_ci <- function(x, level = 0.95) {
  se <- sd(x) / sqrt(length(x))
  alpha <- 1 - level
  mean(x) + se * qnorm(c(alpha / 2, 1 - alpha / 2))
}
```

4. Return statements
One of your colleagues has noticed if you pass `mean_ci()` an empty vector it returns a confidence interval with missing values at both ends (try it: `mean_ci(numeric(0)))`. In this case, they decided it would make more sense to produce a warning `"x was empty"` and return `c(-Inf, Inf)` and have edited the function to be:

```{r}
mean_ci <- function(x, level = 0.95) {
  if (length(x) == 0) {
    warning("x was empty", call. = FALSE)
    return(c(-Inf, Inf))
  } else {
    se <- sd(x) / sqrt(length(x))
    alpha <- 1 - level
    mean(x) + se * qnorm(c(alpha / 2, 1 - alpha / 2))
  }
}
mean_ci(numeric(0))
```

5. Rewrite a poorly-written function

```{r poorly_written_function}
f <- function(x, y) {
  x[is.na(x)] <- y
  cat(sum(is.na(x)), y, "\n")
  x
}
```

```{r test_function}
# Define a numeric vector x with the values 1, 2, NA, 4 and 5
x <- c(1, 2, NA, 4, 5)
# Call f() with the arguments x = x and y = 3
f(x = x, y = 3)
# Call f() with the arguments x = x and y = 10
f(x = x, y = 10)
```

a. Let's make it clear from its name

Did you figure out what `f()` does? `f()` takes a vector `x` and replaces any missing values in it with the value `y`.

Imagine you came across the line `df$z <- f(df$z, 0)` a little further on in the code. What does that line do? Now you know, it replaces any missing values in the column `df$z` with the value 0. But anyone else who comes across that line is going to have to go back and find the definition of `f` and see if they can reason it out.

Let's rename our function and arguments to make it obvious to everyone what we are trying to achieve.

```{r}
df <- data.frame(
  z = c(-0.9403678, 1.3296136, 
        NA, 0.1831795, 
        0.5811088, 1.7948628, 
        NA, -1.8137820, 
        1.4397629, -1.6145781)
)
# Rename the function f() to replace_missings()
replace_missings <- function(x, replacement) {
  # Change the name of the y argument to replacement
  x[is.na(x)] <- replacement
  cat(sum(is.na(x)), replacement, "\n")
  x
}
# Rewrite the call on df$z to match our new names
df$z <- replace_missings(x = df$z, replacement = 0)
dplyr::glimpse(df)
```

b. Make the body more understandable

Great! Now when someone comes across 

> df$z <- replace_missings(df$z, replacement = 0)

in your code, it's pretty obvious what you are trying to achieve. The body of our `replace_missings()` function is still a little messy. There is certainly some unnecessary duplication.

```{r}
replace_missings <- function(x, replacement) {
  # Define is_miss
  is_miss <- is.na(x)
  
  # Rewrite rest of function to refer to is_miss
  x[is_miss] <- replacement
  cat(sum(is_miss), replacement, "\n")
  x
}
```

c. Much better! But a few more tweaks

Did you notice replace_missings() prints some output to the console? That output isn't exactly self-explanatory. It would be much nicer to say `"2 missing values replaced by the value 0"`. 

It is also bad practice to use `cat()` for anything other than a `print()` method (a function designed just to display output). Having an important message just print to the screen makes it very hard for other people who might be programming with your function to capture the output and handle it appropriately. 

The official R way to supply simple diagnostic information is the `message()` function. The unnamed arguments are pasted together with no separator (and no need for a newline at the end) and by default are printed to the screen.

Let's make our function nicer for users by using `message()` and making the output self-contained.

```{r}
replace_missings <- function(x, replacement) {
  is_miss <- is.na(x)
  x[is_miss] <- replacement
  
  # Rewrite to use message()
  message(sum(is_miss), " missing values replaced by the value ", replacement)
  x
}

# Check your new function by running on df$z
df <- data.frame(
  z = c(-0.9403678, 1.3296136, 
        NA, 0.1831795, 
        0.5811088, 1.7948628, 
        NA, -1.8137820, 
        1.4397629, -1.6145781)
)

df$z <- replace_missings(df$z, replacement = 0)
dplyr::glimpse(df)
```

