---
title: "Writing Functions in R (Chapter 3)"
subtitle: "Functional Programming"
author: "Seun Odeyemi"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: united
    highlight: tango
    toc: true
    toc_depth: 4
    toc_float: true
---

```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(error = TRUE, 
                      collapse = TRUE, 
                      comment = "#>")
library(styler)
library(lintr)
library(purrr)
```

# Why Functional Programming? 

In this next section you are going to learn some alternatives to the for loop. If you've use R---even for a little while---I'm sure someone has told you that for loops are bad. (That's definitely not true). But there are some tools that make your life easier by writing the for loops for you. You are not going to be able to do anything you couldn't do before, but you'll be able to do it with greater ease. These are the tools of functional programming.

```{r using_map_in_purrr_package}
means <- map_dbl(mtcars, mean)
medians <- map_dbl(mtcars, median)
means
medians
```

# Practice 1

1. Using a for loop to remove duplication

Imagine we have a dataframe `df`:

```{r }
df <- data.frame(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)
```

We want to compute the median of each column. You could do this with copy and paste.

```{r}
median(df[[1]])
median(df[[2]])
median(df[[3]])
median(df[[4]])
```

But that's a lot of repetition! Let's start by seeing how we could reduce the duplication by using a for loop.

```{r}
# Initialize output vector
output <- numeric(ncol(df))

# Fill in the body of the for loop
for (i in seq_along(df)) {            
output[[i]] <- median(df[[i]], na.rm = TRUE)
}

# View the result
output
```

2. Turning the for loop into a function

Now, imagine you need to do this to another data frame `df2`. You copy and paste the `for` loop, and edit every reference to `df` to be `df2` instead.

```{r turning_loop_to_fx}
output <- numeric(ncol(df)) 
for (i in seq_along(df2)) {            
  output[[i]] <- median(df2[[i]])      
}
output
```

And then you realize you have another data frame `df3` for which you also want the column medians. You copy and paste ... and realize you've copied and pasted two times. Time to write a function!

```{r create_col_median_fx}
# Turn this code into col_median()
col_median <- function(df) {
  # initialize the output vector
  output <- numeric(ncol(df))
  # create for loop
  for (i in seq_along(df)) {            
    output[[i]] <- median(df[[i]])      
}
output
}
col_median(df)
```

3. What about column means?

What if instead of medians of every column you actually want means?

Let's write a `col_mean()` function that returns the vector of column means.

```{r create_col_mean_fx}
# Change col_median() to a col_mean() function to find column means
col_mean <- function(df) {
  # initialize the output vector
  output <- numeric(ncol(df))
  # create for loop
  for (i in seq_along(df)) {            
    output[[i]] <- mean(df[[i]])      
}
output
}
col_mean(df)
```

4. What about column standard deviations?

You now have functions for column medians and means, what about one for standard deviations?

```{r create_col_sd_fx}
# Define col_sd() function
col_sd <- function(df) {
  # initialize the output vector
  output <- numeric(ncol(df))
  # create for loop
  for (i in seq_along(df)) {            
    output[[i]] <- sd(df[[i]])      
}
output
}
col_sd(df)
```

5. Uh oh ... time to write a function again

We just copied and pasted the function `col_median` two times. That's a sure sign we need to write a function. How can we write a function that will take column summaries for any summary function we provide?

Let's look at a simpler example first. Consider the functions `f1()`, `f2()` and `f3()` that take a vector `x` and return deviations from the mean value raised to the powers 1, 2, and 3 respectively:

```{r create_f1_f2_f3}
f1 <- function(x) abs(x - mean(x)) ^ 1
f2 <- function(x) abs(x - mean(x)) ^ 2
f3 <- function(x) abs(x - mean(x)) ^ 3
```

Hopefully, you would suggest writing a single function with two arguments: `x` _and_ `power`. That way, one function reproduces all the functionality of `f1()`, `f2()` and `f3()`, and more.

```{r a_more_generalizable_fx}
x <- c(1:10)
# Add a second argument called power
f <- function(x, power) {
  # Edit the body to return absolute deviations raised to power
  abs(x - mean(x)) ^ power
}
f(x,4)
```

# Using a function as an argument

You just saw that we can remove the duplication in our set of summary functions by requiring the function doing the summary as an input. This leads to creating the `col_summary` function:

```{r col_summary} 
col_summary <- function(df, fun) {
  output <- numeric(ncol(df))
  for (i in seq_along(df)) {
    output[[i]] <- fun(df[[i]])
  }
  output
}
```

It may be kind of surprising that you can pass a function as an argument to another function, so let's verify first that it worked. We've found the column means and medians using our old `col_mean()` and `col_median()` functions. Your job is to repeat the calculations using `col_summary()` instead and verify that it
works.

Not only does `col_summary()` remove the duplication in the functions we've already written, it also allows us to apply any summary to the columns of a data frame. Verify this, by finding the column interquartile ranges using the function `IQR()`. (For more info, see ?IQR.)

```{r}
# Find the column medians using col_median() and col_summary()
col_median(df)
col_summary(df, fun = median)
# Find the column means using col_mean() and col_summary()
col_mean(df)
col_summary(df, fun = mean)
# Find the column IQRs using col_summary()
col_summary(df, IQR)
```

# Introducing purrr

Passing functions as arguments is a very powerful idea. An idea you've seen before in the `lapply()` and `sapply()` functions. For example, we could take the column mean of the dataframe, `df` in a similar way using:

```{r compare_with_apply_functions}
col_summary(df, mean)
lapply(df, mean) # returns a list
sapply(df, mean) # returns a vector
vapply(df, mean, FUN.VALUE = double(1)) # need to vector type for the return value of FUN
```

With `col_summary()` you've essentially written your own version of `sapply()`. Now, it is time to introduce you to a family of functions that perform this kind of iteration: the `map()` functions in the `purrr` package. For example, one of these functions `map_dbl()` works exactly like `sapply()` and our `col_summary()` function. 

```{r using_map_functions}
map_dbl(df, mean)
```

Every map function in the purrr package works the same way. 

```{r eval=FALSE}
map_dbl(.x, .f, ...)
```

1. They take a vector `.x` as a first argument.
2. Do something to each element by applying a function `.f` as a second argument. 
3. The map function loops over the vector, applies a function, and returns a result. 

While all the map functions work the same way, they do differ in their of objects they return. There is one map function for each type of vector.

* `map()`---without a suffix---returns a list.
* `map_dbl()` returns a double vector.
* `map_lgl()` returns a logical vector.
* `map_int()` returns a integer vector.
* `map_chr()` returns a character vector. 

The first argument `.x` is always a vector. But it's worth discussing the slight difference in the action of the map() function depending on the type of vector. So far we've been working with dataframes, if you pass a dataframe to a map function:

```{r}
df_new <- data.frame(a = 1:10, b = 11:20)
map(df_new, mean)
```

`map()` iterates over the columns of the dataframe. Then the function in `.f` is applied to each column in `.x`. Here, map applies mean first to `a` column and then to be `b` column and returns the results as a list. If `.x` is a list, the `map()` function iterates over the elements in the list and returns a list. 

```{r}
l <- list(a = 1:10, b = 11:20)
map(df, mean)
```

Similarly, the `map()` function  applies mean first to the `a` element and then to the `b` element and returns the result as a list. If `.x` is an atomic vector, the map function iterates over each element in the vector. Here, `map()` function  applies mean first to the `a` element---in this case the single number 1---and then to the `b` element---the single number 2---and returns the result as a list. 

```{r}
vec <- c(a = 1, b = 2)
map(vec, mean)
```

# Advantages of the map functions in purrr

The internals of the `map()` function are very similar to the `col_summary()` function we wrote. You might be asking why use these functions in the `purrr` package when we can easily write our own? There are few advantages:

1. The map functions in purrr provides some useful shortcuts for specifying `.f` when it isn't as simple as calling a single function with a single argument. 
2. The purrr functions are also more consistent than the base `sapply()` or `lapply()` functions, which makes them better for programming (Chapter 5). 
3. Once, you master the functions in purrr since you don't have to write your own iteration functions, you'll find that it takes much less time to solve iteration problems. 

# Practice II

1. The map functions

```{r map_fx_practice}
# Load the purrr package
# library(purrr)

# Use map_dbl() to find column means
map_dbl(df, mean)

# Use map_dbl() to column medians
map_dbl(df, median)

# Use map_dbl() to find column standard deviations
map_dbl(df, sd)
```

2. The argument to the map functions

The map functions use the `...` ("dot dot dot") argument to pass along additional arguments to `.f` each time it’s called. For example, we can pass the `trim` argument to the `mean()` function:

```{r eval=FALSE}
map_dbl(df, mean, trim = 0.5)
```

Multiple arguments can be passed along using commas to separate them. For example, we can also pass the `na.rm` argument to `mean()`:

```{r eval=FALSE}
map_dbl(df, mean, trim = 0.5, na.rm = TRUE)
```

> You don't have to specify the arguments by name, but it is good practice!

You may be wondering why the arguments to `map()` are `.x` and `.f` and not `x `and `f`? It's because `.x` and `.f` are very unlikely to be argument names you might pass through the ..., thereby preventing confusion about whether an argument belongs to `map()` or to the function being mapped.

Let's get a bit of practice with this. We'll apply our new knowledge to a subset of the planes data frame available in the `nycflights13` package. Use `map_dbl()` to find the average and 5th percentile of each column in `planes`.

```{r load_nycflights13_glimpse_planes_df, message=FALSE}
library(nycflights13)
library(dplyr)
dplyr::glimpse(planes)
planes <- planes %>% 
  select(year, engines, seats, speed)
```

```{r}
# Find the mean of each column
map_dbl(planes, mean)
# Find the mean of each column, excluding missing values
map_dbl(planes, mean, na.rm = TRUE)
# Find the 5th percentile of each column, excluding missing values
map_dbl(planes, quantile, probs = c(0.05), na.rm = TRUE)
```

Great work! Other nice features of the map functions are that they're implemented in C, which makes them really fast, and that they generally preserve names from their input in the output. 

3. Picking the right map function

Choosing the right map function is important. You can always use map(), which will return a list. However, if you know what type of output you expect, you are better to use the corresponding function. That way, if you expect one thing and get another, you'll know immediately because the map function will return an error.

For example, try running:

```{r eval=FALSE}
map_lgl(df, mean) 
# Error: Can't coerce element 1 from a double to a logical
```

The map functions are what we call `type consistent`. This means you know exactly what type of output to expect regardless of the input. `map_lgl()` either returns either a logical vector or an error. `map_dbl()` returns either a double or an error.

One way to check the output type is to run the corresponding function on the first element. For example, `mean(df[[1]])` returns a single numeric value, suggesting `map_dbl()`.

Let's get some more practice with the map functions. We've created a new data frame, `df3`, in your workspace. Use a map function to find which columns are numeric, the type of each column, and a summary of each column.

```{r picking_the_right_map_fx}
df3 <- data.frame(
  A = rnorm(10),
  B = rep(LETTERS[seq(from = 1, to = 2)], times = 10),
  C = c(1:10),
  D = rnorm(10)
)
# Find the columns that are numeric
map_lgl(df3, is.numeric)
# Find the type of each column
map_chr(df3, typeof)
# Find a summary of each column
map(df3, summary)
```

# Shortcuts in the map function

1. Shortcuts for specifying .f

```{r}
# An anonymous function defined on the fly
map(df, function(x) sum(is.na(x)))
# An anonymous function defined using the formular shortcut
map(df, ~ sum(is.na(.)))
```

2. Shortcuts when .f is [[

Imagine we have this list of results where each element in the list contains both an `a` and `b` element. Nowe, we want to extract all the `a` elements. One wayw will be to write our own anonymous function that takes a single argument `x` and then subsets out the `a` element. Since, this is such a common operation, map provides two shortcuts to achieve the same result. If `.f` is just a string then the element with that name is extracted from each element in `.x`. Similarly, if you provide just an integer to `.f` then the element at that index is extracted from each element of `.x` 


```{r}
list_of_results <- list(
  list(a = 1, b = "A"),
  list(a = 2, b = "C"),
  list(a = 3, b = "D")
)
# an anonymous function
map_dbl(list_of_results, function(x) x[["a"]])
# string subsetting
map_dbl(list_of_results, "a")
# integer subsetting
map_dbl(list_of_results, 1)
```

Practice III

1. Solive a simple problem first

Our goal is to fit a separate linear regression of miles per gallon (`mpg`) against weight (`wt`) for each group of cars in our list of data frames, where each data frame in our list represents a different group. How should we get started?

First, let's confirm the structure of this list of data frames. Then, we'll solve a simpler problem first: fit the regression to the first group of cars.

```{r split_mtcars_by_cyl}
cyl <- split(mtcars, mtcars$cyl)
# Examine the structure of cyl
str(cyl)
# Extract the first element into four_cyls
four_cyls <- cyl[[1]]
# Fit a linear regression of mpg on wt using four_cyls
lm(formula = mpg ~ wt, data = four_cyls)
```

2. Using an anonymous function

Great! We now have a snippet of code that performs the operation we want on one data frame. One option would be to turn this into a function, for example:

```{r}
fit_reg <- function(df) {
  lm(mpg ~ wt, data = df)
}
```

Then pass this function into map():

```{r}
map(cyl, fit_reg)
```

But it seems a bit much to define a function for such a specific model when we only want to do this once. Instead of defining the function in the global environment, we will just use the function anonymously inside our call to `map()`.

What does this mean? Instead of referring to our function by name in `map()`, we define it on the fly in the .f argument to `map()`.

```{r}
# Rewrite to call an anonymous function
map(cyl, function(df) lm(mpg ~ wt, data = df))
```

3. Using a formula

Writing anonymous functions takes a lot of extra key strokes, so `purrr` provides a shortcut that allows you to write an anonymous function as a one-sided formula instead.

In R, a one-sided formula starts with a `~`, followed by an R expression. In purrr's map functions, the R expression can refer to an element of the `.x` argument using the `.` character.

Let's take a look at an example. Imagine, instead of a regression on each data frame in `cyl`, we wanted to know the mean displacement for each data frame. One way to do this would be to use an anonymous function:

```{r}
map_dbl(cyl, function(df) mean(df$disp))
```

To perform the same operation using the formula shortcut, we replace the function definition (`function(df)`) with the `~`, then when we need to refer to the element of `cyl` the function operates on (in this case `df`), we use a `.`.

```{r}
map_dbl(cyl, ~ mean(.$disp))
```

See how much less typing it involves! It also saves you from coming up with an argument name. Can you rewrite our previous anonymous function using this formula shortcut instead?

```{r}
# Rewrite to use the formula shortcut instead
map(cyl, ~ lm(mpg ~ wt, data = .))
```

4. Using a string

There are also some useful shortcuts that come in handy when you want to subset each element of the `.x` argument. If the `.f` argument to a map function is set equal to a string, let's say `"name"`, then purrr extracts the `"name"` element from every element of `.x`.

This is a really common situation you find yourself in when you work with nested lists. For example, if we have a list of where every element contains an `a` and `b` element:

```{r eval=FALSE}
list_of_results <- list(
  list(a = 1, b = "A"), 
  list(a = 2, b = "C"), 
  list(a = 3, b = "D")
)
```

We might want to pull out the a element from every entry. We could do it with the string shortcut like this:

```{r eval=FALSE}
map(list_of_results, "a")
```

Now take our list of regression models

```{r eval=FALSE}
map(cyl, ~ lm(mpg ~ wt, data = .))
```

It might be nice to extract the slope coefficient from each model. You'll do this in a few steps: first fit the models, then get the coefficients from each model using the `coef()` function, then pull out the `wt` estimate using the string shortcut.

```{r}
# Save the result from the previous exercise to the variable models
models <- map(cyl, ~ lm(mpg ~ wt, data = .))
# Use map and coef to get the coefficients for each model: coefs
coefs <- map(models, ~ coef(.))
# Use string shortcut to extract the wt coefficient 
map(coefs, "wt")
```

5. Using a numeric vector

Another useful shortcut for subsetting is to pass a numeric vector as the .f argument. This works just like passing a string but subsets by index rather than name. For example, with your previous `list_of_results`:

```{r}
list_of_results <- list(
  list(a = 1, b = "A"), 
  list(a = 2, b = "C"), 
  list(a = 3, b = "D")
)
```

Another way to pull out the `a` element from each list, is to pull out the first element:

```{r}
map(list_of_results, 1)
```

Let's pull out the slopes from our models again, but this time using numeric subsetting. Also, since we are pulling out a single numeric value from each element, let's use `map_dbl()`.

```{r}
# use map_dbl with the numeric shortcut to pull out the second element
map_dbl(coefs, 2)
```

6. Putting it together with pipes

`purrr` also includes a pipe operator: `%>%`. The pipe operator is another shortcut that saves typing, but also increases readability. The explanation of the pipe operator is quite simple: `x %>% f(y)` is another way of writing `f(x, y)`. That is, the left hand side of the pipe, `x`, becomes the first argument to the function, `f()`, on the right hand side of the pipe.

Take a look at our code to get our list of models:

```{r}
cyl <- split(mtcars, mtcars$cyl) 
map(cyl, ~ lm(mpg ~ wt, data = .))
```

We split the data frame `mtcars` and save it as the variable `cyl`. We then pass `cyl` as the first argument to map to fit the models. We could rewrite this using the pipe operator as:

```{r}
split(mtcars, mtcars$cyl) %>% 
  map(~ lm(mpg ~ wt, data = .))
```

We read this as "split the data frame `mtcars` on `cyl`, then use `map()` on the result."

One of the powerful things about the pipe is we can chain together many operations. Here is our complete code, written with pipes, instead assigning each step to a variable and using it in the next step:

```{r}
mtcars %>% 
  split(mtcars$cyl) %>%
  map(~ lm(mpg ~ wt, data = .)) %>%
  map(coef) %>% 
  map_dbl("wt")
```

```{r}
# Define models (don't change)
models <- mtcars %>% 
  split(mtcars$cyl) %>%
  map(~ lm(mpg ~ wt, data = .)) 

# Rewrite to be a single command using pipes 
models %>% 
  map(summary) %>% 
  map_dbl("r.squared")
```

